Python as a machine learning language: experiences from the scikit-learn project

### 2013-06-16 paper.tex
remove subsections in Implementation

\subsection{Cython for performance}

Our implementation guidelines emphasize on writing short and readable code,
easily maintainable by external contributors, while being as efficient as
possible.  Whenever practicable, algorithms implemented in \sklearn are thus
directly written in pure Python. For intensive numerical computations though,
the library mostly relies on NumPy vector operations, which allows for the code
to remain concise and readable while being efficient. For all algorithms that
cannot be easily and efficiently expressed as NumPy operations, we rely on
Cython \citep{behnel2011cython}  for writing the critical parts and achieving
decent performance and scalability. Cython is a compiled programming language
that extends Python with static typing. It produces efficient C extension
modules that are directly importable from the Python runtime system. Examples of
algorithms written in Cython include stochastic gradient descent for linear
models, some graph-based clustering algorithms or decision trees. To facilitate
the adoption of the library, we also try to limit to a bare minimum the set of
external dependencies. As such, NumPy and SciPy -- the two most common
dependencies in the Python scientific ecosystem -- are actually the only two
required dependencies for the library to fully work. Some modules may rely on
other external libraries, but in such cases we tend to prefer to directly ship
them within the codebase rather than adding new dependencies. In particular, the
codebase includes modified versions of the \textsf{LIBSVM} and
\textsf{LIBLINEAR} libraries \citep{chang2011libsvm,fan2008}, both written in
C{}\verb!++! and wrapped using Cython modules.

A problem arising from the use of Cython, however, is that passing pure Python
functions as argument, while possible, degrades performance significantly, due
to the runtime checks needed for each function invocation. This especially
problematic, for example for kernel methods, which would benefit from accepting
a kernel function as argument. A potential solution to this problem would be for
\sklearn to provide a Cython-level API, thereby alllowing the user to pass
Cython-implemented kernel functions.

\subsection{Parallelism}
\label{sec:parallel}

In many cases, machine learning algorithms can be decomposed into independent
smaller tasks that may each be solved in parallel, thereby reducing the overall
computing times. Some algorithms in \sklearn are implemented as such, and
exploit multi-core architectures.

At this point in time, support for multi-core programming in Python -- at least
in the CPython implementation targeted by scikit-learn -- is however suboptimal.
The interpreter supports multithreading, but due to the \textit{global
interpreter lock}, no two threads in a process may execute Python code at the
same time. It follows that multithreading is not a viable solution for
parallelizing parts of an algorithm, even for embarrassingly parallel problems,
such as fitting and evaluating multiple models in a grid search procedure. The
workaround that we use is to rely on the Python multiprocessing module, which
emulates threads using multiple operating system-level processes. However this
solution is not perfect as it may in turn lead to a large memory overhead since
objects (e.g., input datasets) need to be copied within each process own memory
space. In some cases, it may unfortunately even happen that the overhead becomes
so significant that fully exploiting a multi-core architecture becomes
infeasible due memory constraints. As such, a proper implementation of parallel
machine learning algorithms remain a challenging task for the project.

@@

At a lower-level, a potential point of improvement is parallel processing,
which, as described in Section~\ref{sec:parallel}, is only implemented at a
rather coarse-grained level. Cython built-in support for OpenMP
\citep{dagum1998openmp} remains a viable candidate technology for more fine-
grained multicore support in \sklearn.

>>>>

% \subsection{Cython for performance}

Our implementation guidelines emphasize on writing short and readable code,
easily maintainable by external contributors, while being as efficient as
possible.

Whenever practicable, algorithms implemented in \sklearn are thus
directly written in pure Python. For intensive numerical computations though,
the library mostly relies on NumPy vector operations, which allows for the code
to remain concise, readable and efficient. For all algorithms that cannot be
easily and efficiently expressed as NumPy operations, we rely on Cython
\citep{behnel2011cython}  for writing the critical parts and achieving decent
performance and scalability. Cython is a compiled programming language that
extends Python with static typing. It produces efficient C extension modules
that are directly importable from the Python runtime system. Examples of
algorithms written in Cython include stochastic gradient descent for linear
models, some graph-based clustering algorithms or decision trees.  A problem
arising from the use of Cython, however, is that passing pure Python functions
as argument, while possible, degrades performance significantly due to the
runtime checks needed for each function invocation. This is especially
problematic for example for kernel methods, which would benefit from accepting a
kernel function as argument. A potential solution to this problem would be for
\sklearn to provide a Cython-level API, thereby alllowing the user to pass
Cython-implemented kernel functions.

To facilitate the adoption of the library, we also try to limit to a bare
minimum the set of external dependencies. As such, NumPy and SciPy -- the two
most common dependencies in the Python scientific ecosystem -- are actually the
only two required dependencies for the library to fully work. Some modules may
rely on other external libraries, but in such cases we tend to prefer to
directly ship them within the codebase rather than adding new dependencies. In
particular, the codebase includes modified versions of the \textsf{LIBSVM} and
\textsf{LIBLINEAR} libraries \citep{chang2011libsvm,fan2008}, both written in
C{}\verb!++! and wrapped using Cython modules.

Finally, some algorithms in \sklearn also support multi-core architectures.
Machine learning algorithms can indeed often be decomposed into independent
smaller tasks that may each be solved in parallel, thereby reducing the overall
computing times. At this point in time, support for multi-core programming in
Python -- at least in the CPython implementation targeted by scikit-learn -- is
however suboptimal. The interpreter supports multithreading, but due to the
\textit{global interpreter lock}, no two threads in a process may execute Python
code at the same time. It follows that multithreading is not a viable solution
for parallelizing parts of an algorithm, even for embarrassingly parallel
problems, such as fitting and evaluating multiple models in a grid search
procedure. The workaround that we use is to rely on the Python multiprocessing
module, which emulates threads using multiple operating system- level processes.
However this solution is not perfect as it may in turn lead to a large memory
overhead since objects (e.g., input datasets) need to be copied within each
process own memory space. In some cases, it may unfortunately even happen that
the overhead becomes so significant that fully exploiting a multi- core
architecture becomes infeasible due memory constraints. As such, a proper
implementation of parallel machine learning algorithms remain a challenging task
for the project.

@@

At a lower-level, a potential point of improvement is parallel processing,
which, as described in Section~\ref{sec:implementation}, is only implemented at a
rather coarse-grained level. Cython built-in support for OpenMP
\citep{dagum1998openmp} remains a viable candidate technology for more fine-
grained multicore support in \sklearn.

\subsection*{Acknowledgements}

\textcolor{red}{Mention BSD license, INRIA and stuff.}
